### 2015-04-20 paper.Rmd
Work on paper

---
title:  "The Effect of a Second AIDS Defining Illness on Survival in the MACS"
author: 
- "[Jacob Carey](mailto:jcare15@jhu.edu)"
- "[Matt Murrill](mmurril1@jhmi.edu)"
- "[Kim Kunwha](khkim@jhu.edu)"
date:   "`r Sys.Date()`"
output: pdf_document
abstract: |
  We seek to investigate the effect of a second AIDS defining illness on survival. We included people who are seroprevalent as well as seroconverters. We addressed confounding using propensity scores.
---

```{r chunks, echo=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

We read in the data.
```{r data}
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)

# read in horiz data with lots of covariates
covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv") %>%
           # date of death
    mutate(death.date=as.Date("1960-01-01") + FRSTDTHD,
           # date of aids diagnosis
           aids.date=as.Date("1960-01-01") + FRSTAIDD,
           # date of secondary aids diagnosis
           sec.aids.date=as.Date("1960-01-01") + SECAIDD,
           # what was the first aids diagnosis
           frst.aids.diag=AIDSDIAG,
           # what was the second aids diagnosis
           sec.aids.diag=SECAIDSDIAG,
           # did the person have a sec aids diag
           had.sec=year(sec.aids.date) < 2100,
           # did the person die
           died=year(death.date) < 2100,
           age=as.Date("1960-01-01") + DOB,
           exit=ifelse(!died, as.Date("1994-12-31"), death.date),
           censor.date=as.Date("1970-01-01") + as.integer(exit),
           last.alive=as.Date("1960-01-01") + LASTALID,
           censor.date=ifelse(last.alive < censor.date, last.alive, 
                              censor.date),
           dead.at.censor=(death.date < as.Date("1994-12-31"))) %>%
    select(STUDYID, frst.aids.diag, censor.date, sec.aids.diag,
           dead.at.censor)

# read in visit level data
visits <- read_csv("data/hivvert-data04.csv",
                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
    # put visits in correct 1,2, 3,... form
    mutate(visit=VISIT / 10,
           visit.date=mdy(VDATE),
           haart.era.visit=year(visit.date) <= 1994) %>%

    # join to covariates
    left_join(covariates, by=c("STUDYID")) %>%

    # keep MACS only
    filter(COHORT==1,
           # keep people with first ADI only
           frst.aids.diag != 0,
           # restrict to haart era visits right now
           haart.era.visit == TRUE) %>%
    mutate(censor.date=as.Date("1970-01-01")+censor.date) %>%
    select(id=STUDYID, center=CENTER, visit:dead.at.censor)
```

```{r fig1}
# therapy <- visits %>%
#     group_by(year(visit.date)) %>%
#     summarise(therapy=list(table(THRPY))) %>%
#     unnest(therapy)

theme_set(theme_classic())

visits %>%
    group_by(id) %>%
    summarise(sec.aids=min(sec.aids.diag)) %>%
    mutate(sec.aids=plyr::revalue(as.character(sec.aids),
                                  c("0"="No Aids",
                                    "1"="PCP",
                                    "2"="KS",
                                    "3"="NHL",
                                    "4"="Neuro",
                                    "5"="Bacteria OI",
                                    "6"="Fungal OI",
                                    "7"="Protozoa",
                                    "8"="CMV/HSV",
                                    "9"="Wasting",
                                    "10"="Recurrent PNA",
                                    "11"="Cervical Ca"))) %>%
ggplot(., aes(x=sec.aids)) + geom_histogram() +
    xlab("Type of Secondary Aids Diagnosis") +
    ylab("Count") +
    ggtitle("Figure 1") +
    theme(axis.text.x = element_text(angle=-45))
```

### 2015-04-20 R/paper.Rmd
Move paper.Rmd

--- 2015-04-20 paper.Rmd
@@
                                     "10"="Recurrent PNA",
-                                    "11"="Cervical Ca"))) %>%
+                                    "11"="Cervical Ca")),
+           warn_missing=FALSE) %>%
 ggplot(., aes(x=sec.aids)) + geom_histogram() +
@@
     ggtitle("Figure 1") +
-    theme(axis.text.x = element_text(angle=-45))
+    theme(axis.text.x = element_text(angle=-90))
 ```

### 2015-04-21 R/paper.Rmd
Paste link

--- 2015-04-20 paper.Rmd
@@
 ```{r chunks, echo=FALSE}
-knitr::opts_chunk$set(echo=FALSE)
+knitr::opts_chunk$set(echo=FALSE,
+                      message=FALSE)
 ```
@@
     ggtitle("Figure 1") +
-    theme(axis.text.x = element_text(angle=-45))
+    theme(axis.text.x = element_text(angle=-90)) +
+    scale_x_discrete(limits=k
 ```
+
+```{r naive-survival}
+# http://cran.r-project.org/web/packages/survival/vignettes/timedep.pdf
+```

### 2015-05-03 paper.Rmd
Work on naive PH Model

@@
 ```{r chunks, echo=FALSE}
-knitr::opts_chunk$set(echo=FALSE)
+knitr::opts_chunk$set(echo=FALSE,
+                      message=FALSE,
+                      results='asis')
+```
+
+```{r libs}
+library(readr)
+library(dplyr)
+library(ggplot2)
+library(lubridate)
+library(survival)
+theme_set(theme_classic())
 ```
@@
 ```{r data}
-library(readr)
-library(dplyr)
-library(ggplot2)
-library(lubridate)
-
 # read in horiz data with lots of covariates
-covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv") %>%
+covariates <- read_csv("../data/macswihs_horiz2013_copy-1.csv") %>%
            # date of death
@@
                               censor.date),
+           censor.date=as.Date("1970-01-01")+censor.date,
            dead.at.censor=(death.date < as.Date("1994-12-31"))) %>%
-    select(STUDYID, frst.aids.diag, censor.date, sec.aids.diag,
-           dead.at.censor)
+    select(STUDYID, 
+           aids.date,
+           sec.aids.date,
+           frst.aids.diag, 
+           censor.date, 
+           sec.aids.diag,
+           dead.at.censor, 
+           death.date) %>%
+    filter(year(aids.date) < 2100)
 
@@
            haart.era.visit == TRUE) %>%
-    mutate(censor.date=as.Date("1970-01-01")+censor.date) %>%
     select(id=STUDYID, center=CENTER, visit:dead.at.censor)
@@
 
-theme_set(theme_classic())
-
-visits %>%
+data.plot <- visits %>%
     group_by(id) %>%
@@
                                     "10"="Recurrent PNA",
-                                    "11"="Cervical Ca"))) %>%
-ggplot(., aes(x=sec.aids)) + geom_histogram() +
+                                    "11"="Cervical Ca"),
+                                  warn_missing=FALSE))
+ggplot(data.plot, aes(x=sec.aids)) + 
+    geom_histogram() +
     xlab("Type of Secondary Aids Diagnosis") +
@@
     ggtitle("Figure 1") +
-    theme(axis.text.x = element_text(angle=-45))
+    scale_x_discrete(limits=names(sort(-table(data.plot$sec.aids)))) +
+    theme(axis.text.x = element_text(angle=-90))
 ```
+
+We fit a Cox proportional hazards model to the pre haart era data. We did not allow secondary aids diagnosis to be time varying. 
+```{r naive-survival}
+# http://cran.r-project.org/web/packages/survival/vignettes/timedep.pdf
+# https://stat.ethz.ch/R-manual/R-devel/library/survival/html/coxph.html
+data <- covariates %>%
+    filter(year(aids.date) <= 1994) %>%
+    rowwise() %>%
+    transmute(start=aids.date,
+              stop=min(censor.date, as.Date("1994-12-31")),
+              status=(death.date <= as.Date("1994-12-31")),
+              sec.aids=(sec.aids.diag > 0)) %>%
+    mutate_each(funs(as.numeric), start, stop) %>%
+    filter(stop > start)
+
+naive.model <- coxph(Surv(start, stop, status) ~ sec.aids,
+                     data=data)
+knitr::kable(coef(summary(naive.model)),
+             caption="Naive Model")
+```
+
+```{r strat-surv}
+# strat.model <- coxph(Surv(first.aids.date, stop, dead.at.censor) ~ 
+#                      had.sec.aids + strata(haart.era),
+#                      data=data)
+```
+
+```{r propensity-score}
+# fit logit model (and nonparam?) to calculate prop score
+```
+
+```{r matching}
+```

### 2015-05-03 paper.Rmd
Finish Naive PH model

@@
 # read in horiz data with lots of covariates
-covariates <- read_csv("../data/macswihs_horiz2013_copy-1.csv") %>%
+covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv") %>%
            # date of death
@@
 
-We fit a Cox proportional hazards model to the pre haart era data. We did not allow secondary aids diagnosis to be time varying. 
+We fit a Cox proportional hazards model to the pre haart era data. We allowed secondary aids diagnosis to be time varying. 
 ```{r naive-survival}
@@
               stop=min(censor.date, as.Date("1994-12-31")),
-              status=(death.date <= as.Date("1994-12-31")),
-              sec.aids=(sec.aids.diag > 0)) %>%
-    mutate_each(funs(as.numeric), start, stop) %>%
-    filter(stop > start)
+              sec.aids=sec.aids.date,
+              status=(death.date <= as.Date("1994-12-31"))) %>%
+    mutate_each(funs(as.numeric), start, stop, sec.aids) %>%
+    # there are a few obs coming up where secondary and primary are 
+    # diagnosed at the same time.
+    # let's remove those.
+    filter(stop > start,
+           !is.na(sec.aids))
+
+pre.sec <- data %>%
+    rowwise() %>%
+    mutate(stop=min(stop, sec.aids),
+           sec.aids=0) %>%
+    # remove those with same date diag of first and sec aids
+    filter(start < stop)
+
+post.sec <- data %>%
+    filter(sec.aids < stop) %>%
+    mutate(start=sec.aids,
+           sec.aids=1)
+
+time.varying <- bind_rows(pre.sec, post.sec)
 
 naive.model <- coxph(Surv(start, stop, status) ~ sec.aids,
-                     data=data)
+                     data=time.varying)
+
 knitr::kable(coef(summary(naive.model)),

### 2015-05-04 paper.Rmd
Add lots of confounders to data for model. Do adjusted PH

@@
                     col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
+    # keep MACS only
+    filter(COHORT==1) %>%
+    select(STUDYID) %>%
+    distinct(STUDYID)
+
+covariates <- covariates %>%
+    inner_join(visits)
+```
+
+```{r fig1}
+visits <- read_csv("data/hivvert-data04.csv",
+                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
     # put visits in correct 1,2, 3,... form
@@
     select(id=STUDYID, center=CENTER, visit:dead.at.censor)
-```
 
-```{r fig1}
-# therapy <- visits %>%
-#     group_by(year(visit.date)) %>%
-#     summarise(therapy=list(table(THRPY))) %>%
-#     unnest(therapy)
-
+# only for pre haart!
+# need to do again for all data
 data.plot <- visits %>%
@@
 knitr::kable(coef(summary(naive.model)),
+             caption="Naive Model (All Data)")
+```
+
+```{r all-data}
+# read in horiz data with lots of covariates
+covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv") %>%
+           # date of death
+    mutate(death.date=as.Date("1960-01-01") + FRSTDTHD,
+           # date of aids diagnosis
+           aids.date=as.Date("1960-01-01") + FRSTAIDD,
+           # date of secondary aids diagnosis
+           sec.aids.date=as.Date("1960-01-01") + SECAIDD,
+           # what was the first aids diagnosis
+           frst.aids.diag=AIDSDIAG,
+           # what was the second aids diagnosis
+           sec.aids.diag=SECAIDSDIAG,
+           # did the person have a sec aids diag
+           had.sec=year(sec.aids.date) < 2100,
+           # did the person die
+           died=year(death.date) < 2100,
+           dob=as.Date("1960-01-01") + DOB,
+           age.at.aids=difftime(aids.date, dob, unit="weeks") / 52.25,
+           age.at.aids=as.numeric(age.at.aids),
+           exit=death.date,
+           censor.date=as.Date("1970-01-01") + as.integer(exit),
+           last.alive=as.Date("1960-01-01") + LASTALID,
+           censor.date=ifelse(last.alive < censor.date, last.alive, 
+                              censor.date),
+           censor.date=as.Date("1970-01-01")+censor.date) %>%
+    select(STUDYID, 
+           aids.date,
+           sec.aids.date,
+           frst.aids.diag, 
+           censor.date, 
+           sec.aids.diag,
+           death.date,
+           race=RACE,
+           age.at.aids) %>%
+    filter(year(aids.date) < 2100) %>%
+    mutate(haart.era=(year(aids.date) > 1993))
+
+# what are the important confoudners that we can include for the PH model?
+# what about propensity score?
+
+# what about therapy?
+visits <- read_csv("data/hivvert-data04.csv",
+                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
+    # keep MACS only
+    filter(COHORT == 1) %>% 
+    group_by(STUDYID) %>%
+    summarise(drug.use=sum(POTUSE, CRACKUSE, COCUSE, HERUSE, AMPHUSE, 
+                           POPUSE, na.rm=TRUE) > 1,
+              depression=mean(CESD, na.rm=TRUE) < ifelse(drug.use, 
+                                                         16, 23))
+
+covariates <- covariates %>%
+    mutate(aids.year=year(aids.date),
+           aids.month=month(aids.date))
+
+# cd4 closest to primary AIDS diagnosis
+cd4 <- read_csv("data/hivvert-data04.csv",
+                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
+    filter(COHORT == 1, !is.na(CD4N)) %>%
+    select(STUDYID, cd4=CD4N, VDATE) %>%
+    mutate(VDATE=mdy(VDATE),
+           VDATE=as.Date(VDATE)) %>%
+    inner_join(covariates, by=c("STUDYID"="STUDYID")) %>%
+    select(STUDYID, cd4, VDATE, aids.date) %>%
+    mutate(visit.aids.diff=difftime(VDATE, aids.date)) %>%
+    group_by(STUDYID) %>%
+    filter(visit.aids.diff == min(abs(visit.aids.diff))) %>%
+    select(STUDYID, cd4)
+
+# join in cd4 at first AIDS diagnosis
+summ.visits <- visits %>%
+    left_join(cd4)
+
+# add data together
+covariates <- covariates %>%
+    inner_join(summ.visits) %>%
+    mutate(race=ifelse(race == 1, "white", "non-white"))
+```
+
+We also created a naive model that includes **all** MACS data,
+but does not adjust for any confounders, or the pre / post HAART era.
+```{r all-naive-model}
+data <- covariates %>%
+    rowwise() %>%
+    mutate(start=aids.date,
+           stop=censor.date,
+           sec.aids=sec.aids.date,
+           status=(death.date < as.Date("2100-01-01"))) %>%
+    mutate_each(funs(as.numeric), start, stop, sec.aids) %>%
+    # there are a few obs coming up where secondary and primary are 
+    # diagnosed at the same time.
+    # let's remove those.
+    filter(stop > start,
+           !is.na(sec.aids))
+
+pre.sec <- data %>%
+    rowwise() %>%
+    mutate(stop=min(stop, sec.aids),
+           sec.aids=0) %>%
+    # remove those with same date diag of first and sec aids
+    filter(start < stop)
+
+post.sec <- data %>%
+    filter(sec.aids < stop) %>%
+    mutate(start=sec.aids,
+           sec.aids=1)
+
+time.varying <- bind_rows(pre.sec, post.sec)
+
+naive.model <- coxph(Surv(start, stop, status) ~ sec.aids,
+                     data=time.varying)
+
+knitr::kable(coef(summary(naive.model)),
              caption="Naive Model")
@@
 
-```{r strat-surv}
-# strat.model <- coxph(Surv(first.aids.date, stop, dead.at.censor) ~ 
-#                      had.sec.aids + strata(haart.era),
-#                      data=data)
+Next, we adjusted for several confounders, such as Race, Age at baseline, CD4 at baseline, year of AIDS diagnosis, and drug use.
+
+```{r all-adjusted}
+adjusted.model <- coxph(Surv(start, stop, status) ~ sec.aids +
+                        factor(race) + age.at.aids + drug.use +
+                        aids.year + depression + cd4,
+                        data=time.varying)
 ```
 
-```{r propensity-score}
-# fit logit model (and nonparam?) to calculate prop score
+```{r propensity-score, eval=FALSE}
+prop.score <- glm(sec.aids ~ confounder1 + confounder2,
+                  family=binomial,
+                  data=time.varying)
+# prop.score <- glm(sec.aids ~ status + haart.era,
+#                   family=binomial,
+#                   data=time.varying)
+
+time.varying <- time.varying %>%
+    mutate(propensity=predict(prop.score, type="response"))
 ```
 
-```{r matching}
+```{r matching, eval=FALSE}
+library(MatchIt)
 ```
+
+```{r sophisticated-hazards}
+```
+
+```{r graphs}
+```

### 2015-05-05 paper.Rmd
Fix calculation of CD4 at AIDS

@@
     filter(year(aids.date) < 2100) %>%
-    mutate(haart.era=(year(aids.date) > 1993))
+    mutate(haart.era=(year(aids.date) > 1993),
+           aids.year=year(aids.date),
+           aids.month=month(aids.date),
+           race=ifelse(race == 1, "white", "non-white"))
 
@@
               depression=mean(CESD, na.rm=TRUE) < ifelse(drug.use, 
-                                                         16, 23))
-
-covariates <- covariates %>%
-    mutate(aids.year=year(aids.date),
-           aids.month=month(aids.date))
+                                                         16, 23)) %>%
+    ungroup() %>%
+    # remove the 8 missing depression status
+    filter(!is.na(depression))
 
 # cd4 closest to primary AIDS diagnosis
-cd4 <- read_csv("data/hivvert-data04.csv",
+visits <- read_csv("data/hivvert-data04.csv",
                     col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
@@
     group_by(STUDYID) %>%
-    filter(visit.aids.diff == min(abs(visit.aids.diff))) %>%
-    select(STUDYID, cd4)
-
-# join in cd4 at first AIDS diagnosis
-summ.visits <- visits %>%
-    left_join(cd4)
+    filter(abs(visit.aids.diff) == min(abs(visit.aids.diff))) %>%
+    select(STUDYID, cd4) %>%
+    ungroup() %>%
+    right_join(visits)
 
@@
 covariates <- covariates %>%
-    inner_join(summ.visits) %>%
-    mutate(race=ifelse(race == 1, "white", "non-white"))
+    inner_join(visits) 
 ```
@@
 but does not adjust for any confounders, or the pre / post HAART era.
+
 ```{r all-naive-model}
@@
 
-Next, we adjusted for several confounders, such as Race, Age at baseline, CD4 at baseline, year of AIDS diagnosis, and drug use.
+Next, we adjusted for several confounders, such as Race, Age at baseline, CD4 at AIDS diagnosis, year of AIDS diagnosis, and drug use.
 
@@
                         data=time.varying)
+
+knitr::kable(coef(summary(adjusted.model)),
+             caption="PH Model Adjusted for Confounders")
 ```
@@
 ```{r propensity-score, eval=FALSE}
-prop.score <- glm(sec.aids ~ confounder1 + confounder2,
+prop.score <- glm(sec.aids ~ factor(race) + age.at.aids + drug.use +
+                  aids.year + depression + cd4,
                   family=binomial,
                   data=time.varying)
-# prop.score <- glm(sec.aids ~ status + haart.era,
-#                   family=binomial,
-#                   data=time.varying)
 

### 2015-05-05 paper.Rmd
Add commented out code for matching

@@
 library(MatchIt)
+# ########################################################################
+# ## Q9c Match on propensity score and compare the propensity score 
+# ##	distribution overlap between HIV/HAART and HIV-uninfected before
+# ##	and after matching.
+# ########################################################################
+# 
+# ## We will use the package "MatchIt" to perform matches
+# ##	note: you can have the function matchit actually calculate the propensity score
+# ##		but as we've already calculated cpoe we can have matchit use our already 
+# ##		defined propensity score by the option "distance=dat$cpoe" located in the code below
+# ##	note: we chose a caliper of 0.07, by iteratively going through a search between 0 and .5
+# ##		at a 0.01 step and chose the caliper that maximized the percent balance improvement
+# ##		for distance (e.g. propensity score), age, and bmi.
+# 
+# set.seed(8771)
+# match.set <- matchit(hivhrt~age + bmi,data=dat,distance=dat$cpoe,method="nearest",caliper=.07,discard="both")
+# 
+# ## let's check out the diagnostics
+# 
+# ## numerical check
+# print(sum.match <- summary(match.set,standardize=T))
+# 
+# ## graphical check
+# plot(match.set,type="jitter",interactive=F)
+# plot(match.set,type="hist")
+# plot(sum.match,interactive=F)
+# 
+# matched.dat <- match.data(match.set,"all")
+# 
+# ## how many individuals in matched.dat?
+# nrow(matched.dat)
+# 
+# ## How many events do we have in the matched data?
+# sum(matched.dat$dm)
+# 
+# ########################################################################
+# ## Q10a Get HRs after matching
+# ########################################################################
+# 
+# ph.match <- coxph(Surv(time=t,event=dm)~hivhrt,
+# 				data=matched.dat,method="efron")
+# summary(ph.match)
+# 
+# 
+# 
+# ########################################################################
+# ## Q10b Get HRs after matching with higher caliper
+# ########################################################################
+# 
+# 
+# set.seed(1428)
+# match2.set <- matchit(hivhrt~age + bmi,data=dat,distance=dat$cpoe,method="nearest",caliper=.5,discard="both")
+# 
+# ## let's check out the diagnostics
+# 
+# ## numerical check
+# print(sum.match2 <- summary(match2.set,standardize=T))
+# 
+# ## graphical check
+# plot(match2.set,type="jitter",interactive=F)
+# plot(sum.match2,interactive=F)
+# 
+# matched2.dat <- match.data(match2.set,"all")
+# 
+# ## how many individuals in matched.dat?
+# nrow(matched2.dat)
+# 
+# ##
+# sum(matched2.dat$dm)
+# 
+# ph.match2 <- coxph(Surv(time=t,event=dm)~hivhrt,data=matched2.dat,method="efron")
+# summary(ph.match2)
+# 
+# 
+# 
+# 
+# ########################################################################
+# ## Q10c Get HRs after subclassification
+# ########################################################################
+# 
+# 
+# dat$hivhrtn <- as.numeric(dat$hivhrt)-1
+# sub.set <- matchit(hivhrtn~age + bmi,data=dat,method="subclass",sub.by="all",subclass=5,discard="both")
+# 
+# summary(sub.set,standardize=T)
+# plot(sub.set,type="jitter",interactive=F)
+# plot(summary(sub.set,standardize=T),interactive=F)
+# 
+# sub.dat <- match.data(sub.set,"all")
+# ph.sub <- coxph(Surv(time=t,event=dm)~hivhrtn+strata(subclass),data=sub.dat,method="efron")
+# summary(ph.sub)
 ```

### 2015-05-05 paper.Rmd
First stab at prop scores

--- 2015-05-05 paper.Rmd
@@
 # read in horiz data with lots of covariates
-covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv") %>%
+covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv", 
+                       progress=FALSE) %>%
            # date of death
@@
 # read in visit level data
-visits <- read_csv("data/hivvert-data04.csv",
+visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
                     col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
@@
 ```{r fig1}
-visits <- read_csv("data/hivvert-data04.csv",
+visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
                     col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
@@
 # what about therapy?
-visits <- read_csv("data/hivvert-data04.csv",
+visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
                     col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
@@
 # cd4 closest to primary AIDS diagnosis
-visits <- read_csv("data/hivvert-data04.csv",
+visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
                     col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
@@
 
-```{r propensity-score, eval=FALSE}
-prop.score <- glm(sec.aids ~ factor(race) + age.at.aids + drug.use +
+Finally, we use propensity score matching to account for confounding bias.
+
+```{r propensity-score}
+set.seed(90210)
+match.set <- MatchIt::matchit(sec.aids ~ factor(race) + 
+                              age.at.aids + drug.use +
                               aids.year + depression + cd4,
-                  family=binomial,
-                  data=time.varying)
+                              data=time.varying,
+                              distance="logit",
+                              method="nearest",
+                              caliper=.07,
+                              discard="both")
 
-time.varying <- time.varying %>%
-    mutate(propensity=predict(prop.score, type="response"))
+# ## let's check out the diagnostics
+
+# ## numerical check
+match.check <- summary(match.set, standardize=TRUE)
+
+# ## graphical check
+plot(match.set, type="hist")
+# plot(match.set, type="jitter", interactive=FALSE)
+
+matched.dat <- MatchIt::match.data(match.set,"all")
+
+ph.match <- coxph(Surv(start, stop, status) ~ sec.aids +
+                  factor(race) + age.at.aids + drug.use +
+                  aids.year + depression + cd4,
+   				  data=matched.dat,
+                  method="efron")
 ```
 
-```{r matching, eval=FALSE}
-library(MatchIt)
+There were `r length(unique(matched.dat$STUDYID))` individuals in the 
+matched data set of the orginal `r nrow(data)`. There are 
+`r sum(matched.dat$sec.aids)` events in the matched data set.
+We perform a check on the data and see the following summaries.
+
+```{r}
+knitr::kable(match.check$sum.all,
+             caption="Summary of balance for all data")
+knitr::kable(match.check$sum.matched,
+             caption="Summary of balance for matched data")
+knitr::kable(match.check$reduction,
+             caption="Percent Balance Improvement")
+knitr::kable(match.check$nn,
+             caption="Sample Sizes")
 ```
 
-```{r sophisticated-hazards}
+In recalculating our proportional hazards model using propensity score
+in order to handle confounding bias, we see the following results.
+
+```{r}
+knitr::kable(coef(summary(ph.match)),
+             caption="PH Model (Propensity Scores)")
+```
+
+```{r}
+# ########################################################################
+# ## Q10b Get HRs after matching with higher caliper
+# ########################################################################
+# 
+# 
+# set.seed(1428)
+# match2.set <- matchit(hivhrt~age + bmi,data=dat,distance=dat$cpoe,method="nearest",caliper=.5,discard="both")
+# 
+# ## let's check out the diagnostics
+# 
+# ## numerical check
+# print(sum.match2 <- summary(match2.set,standardize=T))
+# 
+# ## graphical check
+# plot(match2.set,type="jitter",interactive=F)
+# plot(sum.match2,interactive=F)
+# 
+# matched2.dat <- match.data(match2.set,"all")
+# 
+# ## how many individuals in matched.dat?
+# nrow(matched2.dat)
+# 
+# ##
+# sum(matched2.dat$dm)
+# 
+# ph.match2 <- coxph(Surv(time=t,event=dm)~hivhrt,data=matched2.dat,method="efron")
+# summary(ph.match2)
+# 
+# 
+# 
+# 
+# ########################################################################
+# ## Q10c Get HRs after subclassification
+# ########################################################################
+# 
+# 
+# dat$hivhrtn <- as.numeric(dat$hivhrt)-1
+# sub.set <- matchit(hivhrtn~age + bmi,data=dat,method="subclass",sub.by="all",subclass=5,discard="both")
+# 
+# summary(sub.set,standardize=T)
+# plot(sub.set,type="jitter",interactive=F)
+# plot(summary(sub.set,standardize=T),interactive=F)
+# 
+# sub.dat <- match.data(sub.set,"all")
+# ph.sub <- coxph(Surv(time=t,event=dm)~hivhrtn+strata(subclass),data=sub.dat,method="efron")
+# summary(ph.sub)
 ```
@@
 ```{r graphs}
+plot(survfit(ph.match))
 ```

### 2015-05-05 paper.Rmd
First draft of practicum done

@@
 
+adjusted.model.ph <- cox.zph(adjusted.model)
+
 knitr::kable(coef(summary(adjusted.model)),
              caption="PH Model Adjusted for Confounders")
+
+knitr::kable(cox.zph(adjusted.model)$table,
+             caption="PH Assumption Test")
 ```
@@
 ```{r propensity-score}
+caliper <- 0.07
 set.seed(90210)
@@
                               method="nearest",
-                              caliper=.07,
+                              caliper=caliper,
                               discard="both")
@@
 `r sum(matched.dat$sec.aids)` events in the matched data set.
-We perform a check on the data and see the following summaries.
+We perform a check on the data and see the following summaries. We used a
+caliper of `r caliper`.
 
 ```{r}
-knitr::kable(match.check$sum.all,
+knitr::kable(match.check$sum.all[,1:4],
              caption="Summary of balance for all data")
-knitr::kable(match.check$sum.matched,
+knitr::kable(match.check$sum.matched[,1:4],
              caption="Summary of balance for matched data")
@@
              caption="PH Model (Propensity Scores)")
+
+knitr::kable(cox.zph(ph.match)$table,
+             caption="PH Assumption Test")
+caliper <- 0.5
 ```
 
+Next, we performed the same PH model using propensity scores, but with a higher
+caliper (`r caliper`).
+
+```{r propensity-score-2}
+set.seed(90210)
+match.set <- MatchIt::matchit(sec.aids ~ factor(race) + 
+                              age.at.aids + drug.use +
+                              aids.year + depression + cd4,
+                              data=time.varying,
+                              distance="logit",
+                              method="nearest",
+                              caliper=caliper,
+                              discard="both")
+
+# ## let's check out the diagnostics
+
+# ## numerical check
+match.check <- summary(match.set, standardize=TRUE)
+
+# ## graphical check
+plot(match.set, type="hist")
+# plot(match.set, type="jitter", interactive=FALSE)
+
+matched.dat <- MatchIt::match.data(match.set,"all")
+
+ph.match <- coxph(Surv(start, stop, status) ~ sec.aids +
+                  factor(race) + age.at.aids + drug.use +
+                  aids.year + depression + cd4,
+   				  data=matched.dat,
+                  method="efron")
+```
+
+There were `r length(unique(matched.dat$STUDYID))` individuals in the 
+new matched data set of the orginal `r nrow(data)`. There are 
+`r sum(matched.dat$sec.aids)` events in the matched data set.
+We perform a check on the data and see the following summaries.
+
 ```{r}
-# ########################################################################
-# ## Q10b Get HRs after matching with higher caliper
-# ########################################################################
-# 
-# 
-# set.seed(1428)
-# match2.set <- matchit(hivhrt~age + bmi,data=dat,distance=dat$cpoe,method="nearest",caliper=.5,discard="both")
-# 
-# ## let's check out the diagnostics
-# 
-# ## numerical check
-# print(sum.match2 <- summary(match2.set,standardize=T))
-# 
-# ## graphical check
-# plot(match2.set,type="jitter",interactive=F)
-# plot(sum.match2,interactive=F)
-# 
-# matched2.dat <- match.data(match2.set,"all")
-# 
-# ## how many individuals in matched.dat?
-# nrow(matched2.dat)
-# 
-# ##
-# sum(matched2.dat$dm)
-# 
-# ph.match2 <- coxph(Surv(time=t,event=dm)~hivhrt,data=matched2.dat,method="efron")
-# summary(ph.match2)
-# 
-# 
-# 
-# 
-# ########################################################################
-# ## Q10c Get HRs after subclassification
-# ########################################################################
-# 
-# 
-# dat$hivhrtn <- as.numeric(dat$hivhrt)-1
-# sub.set <- matchit(hivhrtn~age + bmi,data=dat,method="subclass",sub.by="all",subclass=5,discard="both")
-# 
+knitr::kable(match.check$sum.all[,1:4],
+             caption="Summary of balance for all data")
+knitr::kable(match.check$sum.matched[,1:4],
+             caption="Summary of balance for matched data")
+knitr::kable(match.check$reduction,
+             caption="Percent Balance Improvement")
+knitr::kable(match.check$nn,
+             caption="Sample Sizes")
+```
+
+In our second PH model with propensity scores, we see the following results.
+
+```{r}
+knitr::kable(coef(summary(ph.match)),
+             caption="PH Model (Propensity Scores)")
+
+knitr::kable(cox.zph(ph.match)$table,
+             caption="PH Assumption Test")
+caliper <- 0.5
+```
+
+Finally, we used a subclassification method for our propensity score.
+
+```{r subclass}
+sub.set <- MatchIt::matchit(sec.aids ~ factor(race) + 
+                            age.at.aids + drug.use +
+                            aids.year + depression + cd4,
+                            data=time.varying,
+                            method="subclass",
+                            sub.by="all",
+                            subclass=5)
+
 # summary(sub.set,standardize=T)
-# plot(sub.set,type="jitter",interactive=F)
+plot(sub.set,type="jitter",interactive=F)
 # plot(summary(sub.set,standardize=T),interactive=F)
-# 
-# sub.dat <- match.data(sub.set,"all")
-# ph.sub <- coxph(Surv(time=t,event=dm)~hivhrtn+strata(subclass),data=sub.dat,method="efron")
-# summary(ph.sub)
+
+sub.dat <- MatchIt::match.data(sub.set, "all")
+ph.match <- coxph(Surv(start, stop, status) ~ sec.aids + strata(subclass),
+   				  data=sub.dat,
+                  method="efron")
+knitr::kable(coef(summary(ph.match)),
+             caption="PH Model (Propensity Scores)")
+
+knitr::kable(cox.zph(ph.match)$table,
+             caption="PH Assumption Test")
 ```
@@
 ```{r graphs}
-plot(survfit(ph.match))
+plot(survfit(ph.match), main="Final PH Model Using Subclassification")
 ```

### 2015-05-08 paper.Rmd
Make notes of Matt's meeting with TA

@@
 
+data %>%
+    filter(year(death.date) < 2100) %>%
+    mutate(time.to.death=difftime(death.date, aids.date, units="weeks") / 4) %>%
+    group_by(frst.aids.diag) %>%
+    summarise(cases=n(),
+              med.to.death=median(time.to.death))
+
 # what about therapy?
@@
     group_by(STUDYID) %>%
-    summarise(drug.use=sum(POTUSE, CRACKUSE, COCUSE, HERUSE, AMPHUSE, 
-                           POPUSE, na.rm=TRUE) > 1,
-              depression=mean(CESD, na.rm=TRUE) < ifelse(drug.use, 
-                                                         16, 23)) %>%
+    summarise(drug.use=sum(CRACKUSE, COCUSE, HERUSE, AMPHUSE, na.rm=TRUE) > 1,
+              depression=mean(CESD, na.rm=TRUE) < 23) %>%
     ungroup() %>%
@@
 ```
+
+Keri's recommendations:
+    1. Origin being primary diagnosis is fine
+    2. person time since prim diag is fine
+    3. be careful w/ term time varying exposure
+    4. covarites: race is fine, age is fine, remove marijauan and poppers from drug
+    5. depression use 23 period.
+    6. should depression be mean?
+    7. CD4 is great
+    8. year of aids diagnosis? restrict to 1996. Censor study at 1996. could
+       categorize year (<90, 90-93, 93-Jul 95, >Jul 95)
+    9. put people back in that had sec aids at baseline.
+    10. timing of secondary aids diagnosis? distribution of prim aids to sec. aids
+        infection?
+    11. sens. look at dist of survival by type. pick a cutoff that we think is
+        meaningful. type of primary. i.e. higher median survival, lower median
+        survival.
+    12. switch propensity scores to ipw
+    13. combinations of prim and sec aids impact. could explore with
+        interaction.
+    14. Add site

### 2015-05-09 paper.Rmd
Many things...

--- 2015-05-05 paper.Rmd
@@
 - "[Jacob Carey](mailto:jcare15@jhu.edu)"
-- "[Matt Murrill](mmurril1@jhmi.edu)"
-- "[Kim Kunwha](khkim@jhu.edu)"
+- "[Matt Murrill](mailto:mmurril1@jhmi.edu)"
+- "[Kim Kunwha](mailto:khkim@jhu.edu)"
 date:   "`r Sys.Date()`"
-output: pdf_document
+output: word_document
 abstract: |
@@
 library(dplyr)
+library(tidyr)
 library(ggplot2)
@@
 
-We read in the data.
-```{r data}
-# read in horiz data with lots of covariates
-covariates <- read_csv("data/macswihs_horiz2013_copy-1.csv", 
-                       progress=FALSE) %>%
-           # date of death
-    mutate(death.date=as.Date("1960-01-01") + FRSTDTHD,
-           # date of aids diagnosis
-           aids.date=as.Date("1960-01-01") + FRSTAIDD,
-           # date of secondary aids diagnosis
-           sec.aids.date=as.Date("1960-01-01") + SECAIDD,
-           # what was the first aids diagnosis
-           frst.aids.diag=AIDSDIAG,
-           # what was the second aids diagnosis
-           sec.aids.diag=SECAIDSDIAG,
-           # did the person have a sec aids diag
-           had.sec=year(sec.aids.date) < 2100,
-           # did the person die
-           died=year(death.date) < 2100,
-           age=as.Date("1960-01-01") + DOB,
-           exit=ifelse(!died, as.Date("1994-12-31"), death.date),
-           censor.date=as.Date("1970-01-01") + as.integer(exit),
-           last.alive=as.Date("1960-01-01") + LASTALID,
-           censor.date=ifelse(last.alive < censor.date, last.alive, 
-                              censor.date),
-           censor.date=as.Date("1970-01-01")+censor.date,
-           dead.at.censor=(death.date < as.Date("1994-12-31"))) %>%
-    select(STUDYID, 
-           aids.date,
-           sec.aids.date,
-           frst.aids.diag, 
-           censor.date, 
-           sec.aids.diag,
-           dead.at.censor, 
-           death.date) %>%
-    filter(year(aids.date) < 2100)
-
-# read in visit level data
-visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
-                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
-    # keep MACS only
-    filter(COHORT==1) %>%
-    select(STUDYID) %>%
-    distinct(STUDYID)
-
-covariates <- covariates %>%
-    inner_join(visits)
-```
-
-```{r fig1}
-visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
-                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
-    # put visits in correct 1,2, 3,... form
-    mutate(visit=VISIT / 10,
-           visit.date=mdy(VDATE),
-           haart.era.visit=year(visit.date) <= 1994) %>%
-
-    # join to covariates
-    left_join(covariates, by=c("STUDYID")) %>%
-
-    # keep MACS only
-    filter(COHORT==1,
-           # keep people with first ADI only
-           frst.aids.diag != 0,
-           # restrict to haart era visits right now
-           haart.era.visit == TRUE) %>%
-    select(id=STUDYID, center=CENTER, visit:dead.at.censor)
-
-# only for pre haart!
-# need to do again for all data
-data.plot <- visits %>%
-    group_by(id) %>%
-    summarise(sec.aids=min(sec.aids.diag)) %>%
-    mutate(sec.aids=plyr::revalue(as.character(sec.aids),
-                                  c("0"="No Aids",
+```{r all-data}
+# bins for 2nd ADI
+# PCP, KS, CMV, NHL, neuro, other
+map <- c("0"="No Aids",
          "1"="PCP",
@@
          "4"="Neuro",
-                                    "5"="Bacteria OI",
-                                    "6"="Fungal OI",
-                                    "7"="Protozoa",
+         "5"="Other",
+         "6"="Other",
+         "7"="Other",
          "8"="CMV/HSV",
-                                    "9"="Wasting",
-                                    "10"="Recurrent PNA",
-                                    "11"="Cervical Ca"),
-                                  warn_missing=FALSE))
-ggplot(data.plot, aes(x=sec.aids)) + 
-    geom_histogram() +
-    xlab("Type of Secondary Aids Diagnosis") +
-    ylab("Count") +
-    ggtitle("Figure 1") +
-    scale_x_discrete(limits=names(sort(-table(data.plot$sec.aids)))) +
-    theme(axis.text.x = element_text(angle=-90))
-```
+         "9"="Other",
+         "10"="Other",
+         "11"="Other")
 
-We fit a Cox proportional hazards model to the pre haart era data. We allowed secondary aids diagnosis to be time varying. 
-```{r naive-survival}
-# http://cran.r-project.org/web/packages/survival/vignettes/timedep.pdf
-# https://stat.ethz.ch/R-manual/R-devel/library/survival/html/coxph.html
-data <- covariates %>%
-    filter(year(aids.date) <= 1994) %>%
-    rowwise() %>%
-    transmute(start=aids.date,
-              stop=min(censor.date, as.Date("1994-12-31")),
-              sec.aids=sec.aids.date,
-              status=(death.date <= as.Date("1994-12-31"))) %>%
-    mutate_each(funs(as.numeric), start, stop, sec.aids) %>%
-    # there are a few obs coming up where secondary and primary are 
-    # diagnosed at the same time.
-    # let's remove those.
-    filter(stop > start,
-           !is.na(sec.aids))
+map.2 <- c("1"="Baltimore",
+           "2"="Chicago",
+           "3"="Pittsburgh",
+           "4"="Los Angeles")
 
-pre.sec <- data %>%
-    rowwise() %>%
-    mutate(stop=min(stop, sec.aids),
-           sec.aids=0) %>%
-    # remove those with same date diag of first and sec aids
-    filter(start < stop)
-
-post.sec <- data %>%
-    filter(sec.aids < stop) %>%
-    mutate(start=sec.aids,
-           sec.aids=1)
-
-time.varying <- bind_rows(pre.sec, post.sec)
-
-naive.model <- coxph(Surv(start, stop, status) ~ sec.aids,
-                     data=time.varying)
-
-knitr::kable(coef(summary(naive.model)),
-             caption="Naive Model (All Data)")
-```
-
-```{r all-data}
 # read in horiz data with lots of covariates
@@
            aids.date=as.Date("1960-01-01") + FRSTAIDD,
+           # aids interval
+           aids.interval=ifelse(year(aids.date) < 1990, "[84, 90)", 
+                         ifelse(year(aids.date) < 1993, "[90, 93)",
+                         ifelse(aids.date < as.Date("1995-07-01"), 
+                                "[93, 07-95)",
+                                "[07-95, 96)"))),
            # date of secondary aids diagnosis
            sec.aids.date=as.Date("1960-01-01") + SECAIDD,
+           # sec aids interval
+           seca.interval=ifelse(year(sec.aids.date) < 1990, "[84, 90)", 
+                         ifelse(year(sec.aids.date) < 1993, "[90, 93)",
+                         ifelse(sec.aids.date < as.Date("1995-07-01"), 
+                                "[93, 07-95)",
+                                "[07-95, 96)"))),
            # what was the first aids diagnosis
-           frst.aids.diag=AIDSDIAG,
-           # what was the second aids diagnosis
-           sec.aids.diag=SECAIDSDIAG,
+           frst.aids.diag=plyr::revalue(as.character(AIDSDIAG), map, 
+                                        warn_missing=FALSE),
+           # what was the first aids diagnosis
+           sec.aids.diag=plyr::revalue(as.character(SECAIDSDIAG), map, 
+                                       warn_missing=FALSE),
            # did the person have a sec aids diag
-           had.sec=year(sec.aids.date) < 2100,
-           # did the person die
-           died=year(death.date) < 2100,
+           had.sec=year(sec.aids.date) < 1996,
            dob=as.Date("1960-01-01") + DOB,
            age.at.aids=difftime(aids.date, dob, unit="weeks") / 52.25,
-           age.at.aids=as.numeric(age.at.aids),
-           exit=death.date,
-           censor.date=as.Date("1970-01-01") + as.integer(exit),
-           last.alive=as.Date("1960-01-01") + LASTALID,
-           censor.date=ifelse(last.alive < censor.date, last.alive, 
-                              censor.date),
-           censor.date=as.Date("1970-01-01")+censor.date) %>%
+           age.at.aids=as.numeric(age.at.aids)) %>%
     select(STUDYID, 
            aids.date,
+           aids.interval,
            sec.aids.date,
+           seca.interval,
+           had.sec,
            frst.aids.diag, 
-           censor.date, 
            sec.aids.diag,
@@
            age.at.aids) %>%
-    filter(year(aids.date) < 2100) %>%
-    mutate(haart.era=(year(aids.date) > 1993),
-           aids.year=year(aids.date),
+    # only keep pts who developed aids before 1996
+    filter(year(aids.date) < 1996) %>%
+    mutate(aids.year=year(aids.date),
            aids.month=month(aids.date),
-           race=ifelse(race == 1, "white", "non-white"))
+           race=ifelse(race == 1, "white", "non-white")) %>%
+    rowwise() %>%
+    mutate(sec.aids.date=min(sec.aids.date, 
+                             as.Date("1995-12-31"))) %>%
+    ungroup() %>%
+    mutate(sec.aids.date=ifelse(had.sec, sec.aids.date, 
+                                as.Date("2100-01-01")),
+           sec.aids.date=as.Date(sec.aids.date, "1970-01-01"),
+           death.date=ifelse(year(death.date) >= 1996,
+                             as.Date("2100-01-01"),
+                             death.date),
+           death.date=as.Date(death.date, "1970-01-01")) %>%
+    rowwise() %>%
+    mutate(censor.date=min(death.date, as.Date("1995-12-31")),
+           censor.date=as.Date(censor.date, "1970-01-01")) %>%
+    ungroup() %>%
+    tbl_df()
 
-# what are the important confoudners that we can include for the PH model?
-# what about propensity score?
+# get CENTER
+visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
+                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
+    filter(COHORT == 1) %>%
+    count(STUDYID, CENTER) %>%
+    group_by(STUDYID) %>%
+    # just consider the most visited center
+    filter(n==max(n)) %>%
+    # there are a few ties so let's just pick the first one
+    distinct(STUDYID) %>%
+    # forget n
+    select(-n) %>%
+    mutate(CENTER=plyr::revalue(as.character(CENTER), 
+                                map.2, warn_missing=FALSE))
 
-# what about therapy?
+# get drug use and depression status
 visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
@@
     group_by(STUDYID) %>%
-    summarise(drug.use=sum(POTUSE, CRACKUSE, COCUSE, HERUSE, AMPHUSE, 
-                           POPUSE, na.rm=TRUE) > 1,
-              depression=mean(CESD, na.rm=TRUE) < ifelse(drug.use, 
-                                                         16, 23)) %>%
+    summarise(drug.use=sum(CRACKUSE, COCUSE, HERUSE, 
+                           AMPHUSE, na.rm=TRUE) > 1,
+              depression=mean(CESD, na.rm=TRUE) > 16) %>%
     ungroup() %>%
     # remove the 8 missing depression status
-    filter(!is.na(depression))
+    filter(!is.na(depression)) %>%
+    # join in center
+    left_join(visits)
 
@@
 covariates <- covariates %>%
-    inner_join(visits) 
+    inner_join(visits) %>%
+    filter(!is.na(cd4))
 ```
 
-We also created a naive model that includes **all** MACS data,
-but does not adjust for any confounders, or the pre / post HAART era.
+```{r graphs-explr}
+data.plot <- covariates %>%
+    mutate(adi=plyr::revalue(as.character(frst.aids.diag),
+                             map, warn_missing=FALSE),
+           sec.adi=plyr::revalue(as.character(sec.aids.diag),
+                                 map, warn_missing=FALSE)) %>%
+    select(adi,
+           sec.adi,
+           aids.date,
+           sec.aids.date,
+           death.date)
+
+ggplot(data.plot, aes(x=adi)) + 
+    geom_histogram() +
+    xlab("Type of Primary Aids Diagnosis") +
+    ylab("Count") +
+    ggtitle("Distibution of Primary ADI's") +
+    scale_x_discrete(limits=names(sort(-table(data.plot$adi)))) +
+    theme(axis.text.x = element_text(angle=-90))
+```
+
+\newpage
+
+red line is mean.
+
+```{r graphs-2}
+data.plot.1 <- data.plot %>%
+    filter(year(death.date) < 2100) %>%
+    mutate(time.to.death=difftime(death.date, aids.date, units="weeks"),
+           time.to.death=as.numeric(time.to.death) / 52.25)
+
+vline <- data.plot.1 %>%
+    group_by(adi) %>%
+    summarise(mean.time=mean(time.to.death),
+              count=n())
+
+ggplot(data.plot.1, aes(x=time.to.death)) +
+    geom_density() +
+    geom_vline(data=vline,
+               aes(xintercept=mean.time),
+               colour="red",
+               linetype="dashed") +
+    xlab("Years from AIDS Diagnosis to Death") +
+    ggtitle("Distribution of Deadliness of ADI") +
+    facet_wrap(~adi)
+
+data.plot.2 <- data.plot %>%
+    filter(year(death.date) < 2100,
+           year(sec.aids.date) < 2100) %>%
+    mutate(time.to.death=difftime(death.date, sec.aids.date, units="weeks"),
+           time.to.death=as.numeric(time.to.death) / 52.25)
+
+vline <- data.plot.2 %>%
+    group_by(sec.adi) %>%
+    summarise(mean.time=mean(time.to.death),
+              count=n())
+
+ggplot(data.plot.2, aes(x=time.to.death)) +
+    geom_density() +
+    geom_vline(data=vline,
+               aes(xintercept=mean.time),
+               colour="red",
+               linetype="dashed") +
+    xlab("Years from Second AIDS Diagnosis to Death") +
+    ggtitle("Distribution of Deadliness of Secondary ADI") +
+    facet_wrap(~sec.adi)
+```
+
+\newpage
+
+We created a naive model that includes incident and prevalent cases
+diagnoses before 1996 and administratively censored them at 1996.
+We did not adjust for any confounders.
 
@@
     mutate_each(funs(as.numeric), start, stop, sec.aids) %>%
-    # there are a few obs coming up where secondary and primary are 
-    # diagnosed at the same time.
-    # let's remove those.
-    filter(stop > start,
-           !is.na(sec.aids))
+    # remove those with NA for sec aids date
+    # and those with aids diag and death on same day
+    filter(!is.na(sec.aids),
+           aids.date != death.date)
 
@@
            sec.aids=0) %>%
-    # remove those with same date diag of first and sec aids
-    filter(start < stop)
+    # remove those with sec aids == start aids
+    # well put them in post sec
+    filter(stop > start)
 
@@
 
+\newpage
+
 Next, we adjusted for several confounders, such as Race, Age at baseline, CD4 at AIDS diagnosis, year of AIDS diagnosis, and drug use.
@@
                         factor(race) + age.at.aids + drug.use +
-                        aids.year + depression + cd4,
+                        depression + cd4 + 
+                        factor(CENTER) + aids.interval,
                         data=time.varying)
@@
 
-Finally, we use propensity score matching to account for confounding bias.
+\newpage
 
-```{r propensity-score}
-caliper <- 0.07
-set.seed(90210)
-match.set <- MatchIt::matchit(sec.aids ~ factor(race) + 
-                              age.at.aids + drug.use +
-                              aids.year + depression + cd4,
+Finally, we use an IPW approach to adjust for confounding bias.
+
+```{r ipw}
+cpoe.model <- glm(sec.aids ~ factor(race) + age.at.aids +
+                  drug.use + depression + cd4 + 
+                  factor(CENTER) + aids.interval, 
                   data=time.varying,
-                              distance="logit",
-                              method="nearest",
-                              caliper=caliper,
-                              discard="both")
+                  family=binomial)
 
-# ## let's check out the diagnostics
+mpoe.model <- glm(sec.aids ~ 1, data=time.varying, family=binomial)
 
-# ## numerical check
-match.check <- summary(match.set, standardize=TRUE)
+time.varying <- time.varying %>%
+    mutate(sec.aids=as.integer(sec.aids),
+           cpoe=fitted(cpoe.model),
+           mpoe=fitted(mpoe.model),
+           w=ifelse(sec.aids == 1, 1 / cpoe, 1 / (1 - cpoe)),
+           sw=ifelse(sec.aids == 1, mpoe / cpoe, (1 - mpoe) / (1 - cpoe)))
 
-# ## graphical check
-plot(match.set, type="hist")
-# plot(match.set, type="jitter", interactive=FALSE)
+data.graph <- time.varying %>%
+    select(w:sw) %>%
+    gather(weight, value, w:sw)
 
-matched.dat <- MatchIt::match.data(match.set,"all")
+ggplot(data.graph, aes(x=value)) +
+    geom_density() +
+    xlab("Weight") +
+    ylab("") +
+    ggtitle("Distribution of weights") +
+    facet_wrap(~weight)
 
-ph.match <- coxph(Surv(start, stop, status) ~ sec.aids +
-                  factor(race) + age.at.aids + drug.use +
-                  aids.year + depression + cd4,
-   				  data=matched.dat,
+ggplot(time.varying, aes(x=factor(sec.aids), y=log(sw))) +
+    geom_boxplot() +
+    xlab("Secondary AIDS Event") +
+    ylab("Log(SW)") +
+    ggtitle("Spread of Log SW by Secondary AIDS Event")
+
+ph.sw <- coxph(Surv(start, stop, status) ~ sec.aids + cluster(STUDYID),
+               data=time.varying,
+               weights=sw ,
                method="efron")
-```
 
-There were `r length(unique(matched.dat$STUDYID))` individuals in the 
-matched data set of the orginal `r nrow(data)`. There are 
-`r sum(matched.dat$sec.aids)` events in the matched data set.
-We perform a check on the data and see the following summaries. We used a
-caliper of `r caliper`.
+knitr::kable(coef(summary(ph.sw)),
+             caption="PH Model Weighted by SW")
 
-```{r}
-knitr::kable(match.check$sum.all[,1:4],
-             caption="Summary of balance for all data")
-knitr::kable(match.check$sum.matched[,1:4],
-             caption="Summary of balance for matched data")
-knitr::kable(match.check$reduction,
-             caption="Percent Balance Improvement")
-knitr::kable(match.check$nn,
-             caption="Sample Sizes")
-```
-
-In recalculating our proportional hazards model using propensity score
-in order to handle confounding bias, we see the following results.
-
-```{r}
-knitr::kable(coef(summary(ph.match)),
-             caption="PH Model (Propensity Scores)")
-
-knitr::kable(cox.zph(ph.match)$table,
-             caption="PH Assumption Test")
-caliper <- 0.5
-```
-
-Next, we performed the same PH model using propensity scores, but with a higher
-caliper (`r caliper`).
-
-```{r propensity-score-2}
-set.seed(90210)
-match.set <- MatchIt::matchit(sec.aids ~ factor(race) + 
-                              age.at.aids + drug.use +
-                              aids.year + depression + cd4,
-                              data=time.varying,
-                              distance="logit",
-                              method="nearest",
-                              caliper=caliper,
-                              discard="both")
-
-# ## let's check out the diagnostics
-
-# ## numerical check
-match.check <- summary(match.set, standardize=TRUE)
-
-# ## graphical check
-plot(match.set, type="hist")
-# plot(match.set, type="jitter", interactive=FALSE)
-
-matched.dat <- MatchIt::match.data(match.set,"all")
-
-ph.match <- coxph(Surv(start, stop, status) ~ sec.aids +
-                  factor(race) + age.at.aids + drug.use +
-                  aids.year + depression + cd4,
-   				  data=matched.dat,
-                  method="efron")
-```
-
-There were `r length(unique(matched.dat$STUDYID))` individuals in the 
-new matched data set of the orginal `r nrow(data)`. There are 
-`r sum(matched.dat$sec.aids)` events in the matched data set.
-We perform a check on the data and see the following summaries.
-
-```{r}
-knitr::kable(match.check$sum.all[,1:4],
-             caption="Summary of balance for all data")
-knitr::kable(match.check$sum.matched[,1:4],
-             caption="Summary of balance for matched data")
-knitr::kable(match.check$reduction,
-             caption="Percent Balance Improvement")
-knitr::kable(match.check$nn,
-             caption="Sample Sizes")
-```
-
-In our second PH model with propensity scores, we see the following results.
-
-```{r}
-knitr::kable(coef(summary(ph.match)),
-             caption="PH Model (Propensity Scores)")
-
-knitr::kable(cox.zph(ph.match)$table,
-             caption="PH Assumption Test")
-caliper <- 0.5
-```
-
-Finally, we used a subclassification method for our propensity score.
-
-```{r subclass}
-sub.set <- MatchIt::matchit(sec.aids ~ factor(race) + 
-                            age.at.aids + drug.use +
-                            aids.year + depression + cd4,
-                            data=time.varying,
-                            method="subclass",
-                            sub.by="all",
-                            subclass=5)
-
-# summary(sub.set,standardize=T)
-plot(sub.set,type="jitter",interactive=F)
-# plot(summary(sub.set,standardize=T),interactive=F)
-
-sub.dat <- MatchIt::match.data(sub.set, "all")
-ph.match <- coxph(Surv(start, stop, status) ~ sec.aids + strata(subclass),
-   				  data=sub.dat,
-                  method="efron")
-knitr::kable(coef(summary(ph.match)),
-             caption="PH Model (Propensity Scores)")
-
-knitr::kable(cox.zph(ph.match)$table,
+knitr::kable(cox.zph(ph.sw)$table,
              caption="PH Assumption Test")
@@
 
-```{r graphs}
-plot(survfit(ph.match), main="Final PH Model Using Subclassification")
+```{r graphs-final}
+# plot(survfit(ph.sw), main="Final PH Model Using IPW")
 ```
+
+Survival in months.
+```{r table1}
+tbl.func <- function(data, col) {
+    row.data <- data %>%
+        group_by_(col) %>%
+        summarise(col=col,
+                  n=n(),
+                  deaths=sum(died),
+                  death.pct=mean(died),
+                  surv.25=quantile(survival.time, 0.25, na.rm=TRUE),
+                  surv.med=quantile(survival.time, 0.5, na.rm=TRUE),
+                  surv.75=quantile(survival.time, 0.75, na.rm=TRUE))
+
+    names(row.data)[1] <- "characteristic"
+
+    return(row.data)
+}
+
+tbl.pre <- data %>%
+    transmute(site=CENTER,
+              race=race,
+              age.ge.40=ifelse(age.at.aids >= 40,
+                               "age >= 40", "age < 40"),
+              aids.year=cut(year(aids.date), c(1984, 1988, 1992, 1995), 
+                                 include.lowest=TRUE),
+              cd4.base=cut(cd4, c(0, 50, 200, Inf), include.lowest=TRUE),
+              initial.aids=frst.aids.diag,
+              second.aids=sec.aids.diag,
+              depressed=ifelse(depression, "depressed", "not depressed"),
+              drug.user=ifelse(drug.use, "Drug User", "Not a Drug User"),
+              site=CENTER,
+              died=year(death.date) < 2100,
+              survival.time=difftime(death.date, aids.date, 
+                                     units="weeks") / 52.25 * 12,
+              survival.time=as.numeric(survival.time),
+              survival.time=ifelse(died, survival.time, NA))
+
+all.data <- tbl.pre %>%
+    summarise(characteristic="all",
+              col="all",
+              n=n(),
+              deaths=sum(died),
+              death.pct=mean(died),
+              surv.25=quantile(survival.time, 0.25, na.rm=TRUE),
+              surv.med=quantile(survival.time, 0.5, na.rm=TRUE),
+              surv.75=quantile(survival.time, 0.75, na.rm=TRUE))
+
+table.1 <- bind_rows(all.data,
+                     tbl.func(tbl.pre, "race"),
+                     tbl.func(tbl.pre, "age.ge.40"),
+                     tbl.func(tbl.pre, "aids.year"),
+                     tbl.func(tbl.pre, "cd4.base"),
+                     tbl.func(tbl.pre, "initial.aids"),
+                     tbl.func(tbl.pre, "second.aids"),
+                     tbl.func(tbl.pre, "depressed"),
+                     tbl.func(tbl.pre, "drug.user"),
+                     tbl.func(tbl.pre, "site"))
+
+knitr::kable(table.1,
+             digits=4,
+             caption="Table 1")
+```
+
+Keri's recommendations:  
+    1. ~~Origin being primary diagnosis is fine~~  
+    2. ~~person time since prim diag is fine~~  
+    3. be careful w/ term time varying exposure  
+    4. ~~covarites: race is fine, age is fine, remove marijauan and poppers from drug~~  
+    5. ~~depression use 23 period~~  
+    6. should depression be mean?  
+    7. ~~CD4 is great~~  
+    8. ~~year of aids diagnosis? restrict to 1996. Censor study at 1996. could categorize year (<90, 90-93, 93-Jul 95, >Jul 95)~~  
+    9. ~~put people back in that had sec aids at baseline.~~  
+    10. timing of secondary aids diagnosis? distribution of prim aids to sec. aids infection?  
+    11. sens. look at dist of survival by type. pick a cutoff that we think is  meaningful. type of primary. i.e. higher median survival, lower median survival.  
+    12. ~~switch propensity scores to ipw~~  
+    13. combinations of prim and sec aids impact. could explore with interaction.  
+    14. ~~Add site~~

### 2015-05-10 paper.Rmd
Update Table 1

@@
     tbl_df()
+
+visits <- read_csv("data/hivvert-data04.csv", progress=FALSE,
+                    col_types="iiciiididdiiiiiiiddiiiiiiiiiiiidc") %>%
+    filter(COHORT == 1, 
+           year(mdy(VDATE)) < 1996) %>%
+    mutate(year=year(mdy(VDATE))) %>%
+    group_by(year) %>%
+    summarise(n=n(),
+              t=sum(!is.na(THRPY)),
+              pct=t / n)
 
@@
 ```{r table1}
-tbl.func <- function(data, col) {
+tbl.func <- function(data, col, sec=FALSE) {
     row.data <- data %>%
@@
                   surv.med=quantile(survival.time, 0.5, na.rm=TRUE),
-                  surv.75=quantile(survival.time, 0.75, na.rm=TRUE))
+                  surv.75=quantile(survival.time, 0.75, na.rm=TRUE),
+                  surv.25.2=quantile(survival.time.2, 0.25, na.rm=TRUE),
+                  surv.med.2=quantile(survival.time.2, 0.5, na.rm=TRUE),
+                  surv.75.2=quantile(survival.time.2, 0.75, na.rm=TRUE))
+
+    if (sec) {
+        row.data <- row.data %>%
+            select(-surv.25:-surv.75) %>%
+            rename(surv.25=surv.25.2,
+                   surv.med=surv.med.2,
+                   surv.75=surv.75.2) %>%
+            filter(second.aids != "No Aids")
+    } else {
+        row.data <- row.data %>%
+            select(-surv.25.2:-surv.75.2)
+    }
 
@@
               survival.time=as.numeric(survival.time),
-              survival.time=ifelse(died, survival.time, NA))
+              survival.time=ifelse(died, survival.time, NA),
+              survival.time.2=difftime(death.date, sec.aids.date,
+                                       units="weeks") / 52.25 * 12,
+              survival.time.2=as.numeric(survival.time.2),
+              survival.time.2=ifelse(died & had.sec, survival.time.2, NA))
 
@@
                      tbl.func(tbl.pre, "initial.aids"),
-                     tbl.func(tbl.pre, "second.aids"),
+                     tbl.func(tbl.pre, "second.aids", sec=TRUE),
                      tbl.func(tbl.pre, "depressed"),
@@
              caption="Table 1")
+```
+
+```{r sens}
+# do cd4 closest to 2nd aids diag instead
+
 ```

### 2015-05-10 paper.Rmd
Fix intervals

@@
               aids.year=cut(year(aids.date), c(1984, 1988, 1992, 1995), 
-                                 include.lowest=TRUE),
-              cd4.base=cut(cd4, c(0, 50, 200, Inf), include.lowest=TRUE),
+                                 include.lowest=TRUE, right=FALSE),
+              cd4.base=cut(cd4, c(0, 50, 200, Inf), include.lowest=TRUE,
+                           right=FALSE),
               initial.aids=frst.aids.diag,

### 2015-05-10 paper.Rmd
Fix intervals again

--- 2015-05-10 paper.Rmd
@@
            # aids interval
-           aids.interval=ifelse(year(aids.date) < 1990, "[84, 90)", 
-                         ifelse(year(aids.date) < 1993, "[90, 93)",
-                         ifelse(aids.date < as.Date("1995-07-01"), 
-                                "[93, 07-95)",
-                                "[07-95, 96)"))),
+#               aids.year=cut(year(aids.date), c(1984, 1988, 1992, 1995), 
+#                                  include.lowest=TRUE, right=FALSE),
+           aids.interval=cut(year(aids.date), c(1984, 1990, 1993, 1995),
+                             include.lowest=TRUE, right=FALSE),
            # date of secondary aids diagnosis
@@
            # sec aids interval
-           seca.interval=ifelse(year(sec.aids.date) < 1990, "[84, 90)", 
-                         ifelse(year(sec.aids.date) < 1993, "[90, 93)",
-                         ifelse(sec.aids.date < as.Date("1995-07-01"), 
-                                "[93, 07-95)",
-                                "[07-95, 96)"))),
+           seca.interval=cut(year(sec.aids.date), c(1984, 1990, 1993, 1995),
+                             include.lowest=TRUE, right=FALSE),
            # what was the first aids diagnosis
@@
                                "age >= 40", "age < 40"),
-              aids.year=cut(year(aids.date), c(1984, 1988, 1992, 1995), 
-                                 include.lowest=TRUE),
-              cd4.base=cut(cd4, c(0, 50, 200, Inf), include.lowest=TRUE),
+              aids.year=aids.interval,
+#               aids.year=cut(year(aids.date), c(1984, 1988, 1992, 1995), 
+#                                  include.lowest=TRUE, right=FALSE),
+              cd4.base=cut(cd4, c(0, 50, 200, Inf), include.lowest=TRUE,
+                           right=FALSE),
               initial.aids=frst.aids.diag,
